---
title: "Project 7 - Intro to Longitudinal Data"
output: html_notebook
---

## step 1: make a person-period file

```{r}
library(readr)
dat <- read.csv("../classroom.csv")
cc.data <- dat[(complete.cases(dat)),]
cc.data$math0 <- cc.data$mathkind
cc.data$math1 <- cc.data$mathkind + cc.data$mathgain
```

```{r}
firstpart <- cc.data
firstpart$year <- rep(0,1081)
firstpart <- firstpart[,-14]
names(firstpart)[13] <- c("math")

secondpart <- cc.data
secondpart$year <- rep(1,1081)
secondpart <- secondpart[,-13]
names(secondpart)[13] <- c("math")
```

```{r}
library(dplyr)
comb.data <- bind_rows(firstpart, secondpart)
head(comb.data)
```


## Fit a baseline model (called unconditional growth) for this nested longitudinal data, ignoring classrooms:

Model:
$$
MATH = b_0 + b_1YEAR_{ijk} +  \zeta_{0k} + \zeta_{1k}+ \eta_{jk} + \varepsilon_{ijk}
$$
With $\zeta_{k}$ ~ N(0,$\sigma^2_{\zeta}$), $\eta_{jk}$ ~ N(0,$\sigma^2_{\eta}$), and $\varepsilon_{ijk}$ ~ N(0,$\sigma^2_{\varepsilon}$), $corr(\zeta_{0k},\zeta_{1k}) = \rho_{\zeta_{0k}\zeta_{1k}}$ and all other pairs independent of one another.


```{r}
require(lme4)
require(lmerTest)
und.grow <- lmer(math ~ year + (year|schoolid) + (1|schoolid:childid), data=comb.data)
summary(und.grow)
```




## Add student, classroom and school-level fixed effects ; fit the model

Model:
$$
MATH = b_0 + b_1YEAR_{ijk} + b_2HOUSEPOV_{k} + b_3MATHKNOW_{jk} + b_4MATHPREP_{jk} + b_5YEARSTEA_{jk} + b_6SEX_{ijk} + b_7MINORITY_{ijk} + b_7SES_{ijk} \zeta_{0k} + \zeta_{1k}+ \eta_{jk} + \varepsilon_{ijk}
$$

With $\zeta_{k}$ ~ N(0,$\sigma^2_{\zeta}$), $\eta_{jk}$ ~ N(0,$\sigma^2_{\eta}$), and $\varepsilon_{ijk}$ ~ N(0,$\sigma^2_{\varepsilon}$), $corr(\zeta_{0k},\zeta_{1k}) = \rho_{\zeta_{0k}\zeta_{1k}}$ and all other pairs independent of one another.

```{r}
und.grow2 <- lmer(math ~ housepov + mathprep + yearstea + mathknow + sex + minority + ses  + year + (year|schoolid) + (1|schoolid:childid), data=comb.data)
summary(und.grow2)
```


## For year==0: what percent of between school differences were explained as you go from the baseline to the second model?

(373.5-249.2)/373.5 = 0.3327979 ~ 33%

###What percent of between child differences were explained as you go from the baseline to the second model?

(749.0-689.5)/749.0 = 0.07943925 ~ 8%

## For year==1: what percent of between school differences were explained as you go from the baseline to the second model?

((373.5+112.4) - (249.2+114.2))/(373.5+112.4) = 0.2521095 ~ 25%

### What percent of between child differences were explained as you go from the baseline to the second model?

((749.0+112.4) - (689.5+114.2))/(749.0+112.4) = 0.06698398 ~ 7%


## Based on significance...

###  what factors seem useful in describing ("explaining") differences between student outcomes? 

Minority and SES

### Point out the direction of the effect.

Minority (negative)
SES (positive)


## Add a random slope for ses, indep. of the year or intercept but that varies at the school-level 

```{r}
und.grow3 <- lmer(math ~ housepov + mathprep + yearstea + mathknow + sex + minority + ses+(0+ses|schoolid)+ year + (year|schoolid) + (1|schoolid:childid), data=comb.data)
summary(und.grow3)
```


### Is the estimated s.d. (square root of variance) of the random slope associated with SES large enough so that a value +/- 1 s.d. is sufficient to "cancel" (or flip the sign) the fixed effect for this predictor?

The estimated s.d. (6.752) of the random slope associated with SES is not large enough to flip the sign of the fixed effect for this predictor.

### Compute residuals in this final model.  Generate a qq plot and density.


###Is there any reason to question the normality assumption?
After looking at the tails, there is reason to question the normality assumption. 
```{r}
pred.yhat <- predict(und.grow3, re.form = ~0)
res1 <- comb.data$math - pred.yhat
qqnorm(res1)
qqline(res1)
```


##compute BLUPs for all 4 random effects.  generate an all pairs scatter plot matrix (4x4) of these 
