---
title: "R Notebook"
output:
  html_notebook: default
  word_document: default
---

#use "classroom.dta"
#intro to nested longitudinal data
#make a person-period file
#gen math0 = mathkind
#gen math1 = mathkind+mathgain
```{r}
dat <- read.csv("../classroom.csv")
cc.data <- dat[(complete.cases(dat)),]
cc.data$math0 <- cc.data$mathkind
cc.data$math1 <- cc.data$mathkind + cc.data$mathgain
```

#reshape long math, i(childid) j(year)
#save "class_pp", replace

```{r}
class_pp <- reshape(cc.data, varying = c("math0", "math1"), v.names = "math", timevar = "year",
                    times = c(0, 1), direction = "long")

attach(class_pp)
```
#fit a baseline model (called unconditional growth) for this nested longitudinal data, ignoring classrooms:
#mixed math year || schoolid: year, cov(un) || childid:
```{r}
require(lme4)
require(lmerTest)
und.grow <- lmer(math ~ year + (year || schoolid) + (1 | childid), data=class_pp)
summary(und.grow)
```
write down this model
$$MATH_{tijk} = b_0 + b_1YEAR_{tijk} + \zeta_{0k} + \zeta_{1k}YEAR_{tijk}+ \delta_{0ijk} + \varepsilon_{tijk} \textrm{ with } \zeta_{0k} \textrm{~ N(0,} \sigma^2_{\zeta} \textrm{), }\zeta_{1k} \textrm{~ N(0,} \sigma^2_{\zeta} \textrm{), } \delta_{0ijk} \textrm{~ N(0,} \sigma^2_{\delta} \textrm{), and } \varepsilon_{tijk} \textrm{ ~ N(0,} \sigma^2_{\varepsilon} \textrm{) all independent of each other.}$$

*add student, classroom and school-level fixed effects ; fit the model
*write down the model
```{r}
und.grow2 <- lmer(math ~ housepov + mathprep + yearstea + mathknow + sex + minority + ses  + year + (year||schoolid) + (1|schoolid:childid), data=class_pp)
summary(und.grow2)
```
$$MATH_{tijk} = b_0 + b_1YEAR_{tijk} + b_2housepov_{k} + b_3mathprep_{jk} + b_4yearstea_{jk} + b_5mathknow_{jk} + b_6sex_{ijk} + b_7minority_{ijk} + b_8ses_{ijk} \\ + \zeta_{0k} + \zeta_{1k}YEAR_{tijk} + \delta_{0ijk} + \varepsilon_{tijk} \textrm{ with } \zeta_{0k} \textrm{~ N(0,} \sigma^2_{\zeta} \textrm{), }\zeta_{1k} \textrm{~ N(0,} \sigma^2_{\zeta} \textrm{), } \delta_{0ijk} \textrm{~ N(0,} \sigma^2_{\delta} \textrm{), and } \varepsilon_{tijk} \textrm{ ~ N(0,} \sigma^2_{\varepsilon} \textrm{) all independent of each other.}$$
*for year==0:
*what percent of between school differences were explained as you go from the baseline to the second model?
(324.79-196.00)/324.79 = 0.3965331 ~ 40%

*what percent of between child differences were explained as you go from the baseline to the second model?
(725.13-686.14)/725.13 = 0.05376967 ~ 5%

*for year==1:
*what percent of between school differences were explained as you go from the baseline to the second model?
((324.79+88.67) - (196.00+82.13))/(324.79+88.67) = 0.327311 ~ 33%

*what percent of between child differences were explained as you go from the baseline to the second model?
(725.13-686.14)/725.13 = 0.05376967 ~ 5%

*based on significance, 
*  what factors seem useful in describing ("explaining") differences between student outcomes? 
Minority and SES
*    Point out the direction of the effect.
Minority (negative)
SES (positive)

*add a random slope for ses, indep. of the year or intercept but that varies at the school-level 
*   (hint: in STATA, you are adding || schoolid: ses BEFORE THE OTHER || schoolid: year, cov(un) entry.

```{r}
und.grow3 <- lmer(math ~ housepov + mathprep + yearstea + mathknow + sex + minority + ses+(0+ses|schoolid)+ year + (year||schoolid) + (1|schoolid:childid), data=class_pp)
summary(und.grow3)
```


*is the estimated s.d. (square root of variance) of the random slope associated with SES large enough 
*  so that a value +/- 1 s.d. is sufficient to "cancel" (or flip the sign) the fixed effect for this predictor?
The estimated s.d. (6.600) of the random slope associated with SES is not large enough to flip the sign of the fixed effect for this predictor.

*compute residuals in this final model.  generate a qq plot and density (STATA: qnorm; kdensity ..., normal) 
*Is there any reason to question the normality assumption?
After looking at the tails, there is reason to question the normality assumption. 
```{r}
pred.yhat <- predict(und.grow3, re.form = ~0)
res1 <- class_pp$math - pred.yhat
qqnorm(res1)
qqline(res1)
```


*compute BLUPs for all 4 random effects.  generate an all pairs scatter plot matrix (4x4) of these 
*  (STATA: graph matrix ...; R: pairs(...))
```{r}
ranefs <- ranef(und.grow3)
zeta0 <- ranefs$schoolid[, 1]
zeta1 <- ranefs$schoolid[, 2]
zeta3 <- ranefs$schoolid[, 3]
delta0 <-ranefs$`schoolid:childid`[,1]

idx.sch <- match(class_pp$schoolid, sort(unique(class_pp$schoolid))) #locations of the unique indices
idx.child <- match(class_pp$childid, sort(unique(class_pp$childid)))

class_pp$zeta0 <- zeta0[idx.sch]
class_pp$zeta1 <- zeta1[idx.sch]
class_pp$zeta3 <- zeta3[idx.sch]
class_pp$delta0 <- delta0[idx.sch]


pairs(class_pp[,16:19])
```
* note whether or not you identify any concerns from these scatterplots.
Scales are not the same so it's hard to tell if there is anything to be concerned about.