---
title: "R Notebook"
output:
  word_document: default
  html_notebook: default
---

```{r}
library(readr)
dat <- read.csv("../classroom.csv")
cc.data <- dat[(complete.cases(dat)),]
```

*we have two models to consider, but we first focus on the simpler one:
```{r}
attach(cc.data)
cc.data$math1st <- mathkind + mathgain
```

```{r}
#require(lme4)
#require(lmerTest)
fit1 <- lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (1|schoolid/classid), data=cc.data)
summary(fit1)
```

*manually construct the residual that removes only the 'fixed effects'
*hint: predict yhat, xb will generate the prediction for the outcome based on the fixed effects only
* then subtract it from the outcome; call this residual: resFE
```{r}
pred.yhat <- predict(fit1, re.form = ~0) #This ignore random effects
resFE <- cc.data$math1st - pred.yhat 
```

*show that this residual is not independent within schools in some manner.

Below, we have sorted each of the fixed effect residuals for the schools by their median from lowest to highest. If we are assuming independence of residuals withink schools, then, we would expect to see a flat-line showing that residuals across schools is random. However, since we are seeing a positive trending line, we can conclude that residuals are not random and it may be dependent on school.
```{r}
ord<-order(unlist(tapply(resFE,schoolid,median)))
boxplot(split(resFE,schoolid)[ord], main = "resFE 1st BoxPlot")
```


*construct the residual that utilizes the BLUPs for the random effects.  Do it in these stages:
* i) predict and save zeta0 
* ii) predit and save eta0
* iii) generate a new residual, called resFE_RE which subtracts yhat, zeta0 and eta0 from the outcome
*note: there is an easier way to get the residuals in this case, predict ..., residuals, but we need to do it manually.
```{r}
idx.sch <- match(cc.data$schoolid, sort(unique(cc.data$schoolid)))
idx.cls <- match(cc.data$classid, sort(unique(cc.data$classid))) 
```

```{r}
attach(cc.data)
ranefs <- ranef(fit1) #Pulling out the random effects
zeta0 <- ranefs$schoolid[, 1] #Random Intercept for School
eta0 <- ranefs$classid[, 1] #Random Intercept for Class
cc.data$zeta0 <- zeta0[idx.sch] # Putting Random Intercept of School back into the data set
cc.data$eta0 <- eta0[idx.cls] #Putting Random Intercept of Class back into the data set
resFE_RE <- math1st - pred.yhat - cc.data$zeta0 - cc.data$eta0 #This accounts for removing both fixed and random effects and the random effects are removed manually. Therefore, residuals should be less dependent than the previous one. Thus, we expect to see a flatter line.
```


* show that these new residuals, resFE_RE are MUCH LESS (if not completely un-) correlated within school 
* using the same method as before (boxplot?)
```{r}
#We are expecting a flatter line but, the line is upward sloping. Come back to it as a group.
ord2<-order(unlist(tapply(resFE_RE,schoolid,median)))
boxplot(split(resFE_RE,schoolid)[ord2], main = "resFE_RE_BoxPlot 2")
```

### Did we tag?
*examine the two sets of BLUPs (for random effects zeta0 and eta0) for normality
*first 'tag' a single value from each grouping (e.g., school or classroom) so that you only have 
* as many BLUPs as the grouping factor (should be less of an issue in R)


```{r}
plot(density(zeta0))
qqnorm(zeta0)
qqline(zeta0)
```

```{r}
plot(density(eta0))
qqnorm(eta0)
qqline(eta0)
```


### Jemar & Zarni
*if In a simpler setting, with no classroom level, if H0: \sigma^2_{\zeta_0}=0 were true, 
* and you sampled 100 schools of size 10, what would be a potential expected estimate of \sigma_{\zeta_0} 
* if you know that \sigma_{\epsilon}=1 ?  HINT: the Central Limit Theorem applies.  Simulate to find out.
```{r}

```


*now reload the data and fit a slightly more complicated model:
use "classroom.dta"
gen math1st = mathkind + mathgain
```{r}
dat <- read.csv("../classroom.csv")
dat$math1st <- dat$mathkind + dat$mathgain
dat <- dat[(complete.cases(dat)),]
attach(dat)
```



mixed math1st housepov yearstea mathprep mathknow ses sex minority |||schoolid: minority, cov(un) || classid:, reml 
est store M2

#Come back and check (0+minority|schoolid) later
```{r}
#The model is slightly more complicated with an uncorrelated random slope school level variability of minority.
M2 <- lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (0+minority|schoolid) + (1|schoolid) + (1|classid))
print(summary(M2))
```

```{r}
idx.sch <- match(dat$schoolid, sort(unique(dat$schoolid)))
idx.cls <- match(dat$classid, sort(unique(dat$classid))) 
```


*manually construct the residual that removes only the 'fixed effects', call this residual: resFE
```{r}
pred.yhat.2 <- predict(M2, re.form = ~0)
ranefs <- ranef(M2)
zeta0 <- ranefs$schoolid[, 1]
dat$zeta0 <- zeta0[idx.sch]
resFE.2 <- dat$math1st - pred.yhat.2 
```


### Jemar & Zarni

Below, we have sorted each of the fixed effect residuals for the schools by their median from lowest to highest. If we are assuming independence of residuals within schools, then, we would expect to see a flat-line showing that residuals across schools is random. However, since we are seeing a positive trending line which is similar to the one above as the random slope for miniority is non existent with variance being zero, we can conclude that residuals are not random and it appears to be dependent on school like the results of the less complicated model above.

```{r}
ord3<-order(unlist(tapply(resFE.2,schoolid,median)))
boxplot(split(resFE.2,schoolid)[ord3], main = "resFE.2 Boxplot_3")
```

*construct the residual that utilizes the BLUPs for the random effects.  Do it in these stages:
* i) predict and save zeta0 AND zeta1 (you need to give them in reverse order in STATA - ask me why if you want)
* ii) predit and save eta0
* iii) generate a new residual, called resFE_RE which subtracts yhat, zeta0, MINORITY*zeta1 and eta0 
```{r}
ranerr <- ranef(M2)
zeta0 <- ranerr$schoolid[,1]
zeta1 <- ranerr$schoolid[,2] #It is the random slope of the minority intercept
eta0 <- ranerr$classid[,1] #This is the random intercept for the classroom level.
dat$zeta0 <- zeta0[idx.sch]
dat$zeta1 <- zeta1[idx.sch]
dat$eta0 <- eta0[idx.cls]
resFE_RE.2 <- dat$math1st - (pred.yhat.2 + dat$zeta0 + (dat$minority * dat$zeta1) + dat$eta0)#This accounts for removing both fixed and random effects and the random effects are removed manually. Therefore, residuals should be less dependent than the previous one. Thus, we expect to see a flatter line.
```


*      from the outcome

### Jemar & Zarni
## We can check this by subtracting the total sum off and the one which is greater will have a positive total slope? #Shoot an email to Marc on this

* show that these new residuals, resFE_RE are MUCH LESS (if not completely un-) correlated within school 
* using the same method as before (boxplot?)
```{r}
ord4<-order(unlist(tapply(resFE_RE.2,schoolid,median)))
boxplot(split(resFE_RE.2,schoolid)[ord4], main = "resFE_RE.2 Boxlot 4")
```

*examine the three sets of BLUPs (for random effects zeta0 and eta0) for normality

This is the BLUP for random intercept at the school level and for the most part, it appears to be normal except in the tails end.
```{r}
qqnorm(zeta0)
qqline(zeta0)
```

```{r}
#Currently, we have found a random slope of zero for the minority. We will have to come back to this.
qqnorm(zeta1)
qqline(zeta1)
```

The classroom level random intercept appears to be normal except in the tails end.
```{r}
qqnorm(eta0)
qqline(eta0)
```

### Jemar & Zarni
*first 'tag' a single value from each grouping (e.g., school or classroom) so that you only have 
* as many BLUPs as the grouping factor (should be less of an issue in R)

```{r}
require(lattice)
xyplot(mathkind ~ ses | schoolid, data = dat[dat$schoolid < 13, ], type = c("p",
"r")) #close enough
```





*plot zeta0 vs. zeta1 to see whether the estimated correlation is consistent with the observed. 
* Use tag to subset as before


```{r}
plot(jitter(zeta0), zeta1)
text(zeta0, zeta1, labels=schoolid, cex= 0.7, pos=3)

plot(jitter(zeta0[1:10], zeta1[1:10]))
textxy(zeta0[1:10], zeta1[1:10], schoolid[1:10])

```

### Jemar & Zarni Much easier to figure out
*track down those odd points in the scatterplot.  What schools are they?
```{r}


```