---
title: "R Notebook"
output: html_notebook
---

```{r}
library(readr)
dat <- read_csv("classroom.csv")
cc.data <- dat[(complete.cases(dat)),]
```

*we have two models to consider, but we first focus on the simpler one:
```{r}
attach(cc.data)
cc.data$math1st <- mathkind + mathgain
```

```{r}
require(lme4)
require(lmerTest)
fit1 <- lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (1|schoolid/classid), data=cc.data)
summary(fit1)
```

*manually construct the residual that removes only the 'fixed effects'
*hint: predict yhat, xb will generate the prediction for the outcome based on the fixed effects only
* then subtract it from the outcome; call this residual: resFE
```{r}
pred.yhat <- predict(fit1, re.form = ~0)
res1 <- cc.data$math1st - pred.yhat
```

*show that this residual is not independent within schools in some manner.  Hint: boxplot?
```{r}
dec.schoolid <- as.numeric(names(sort(table(schoolid))))
cc.data$schoolid <- ordered(cc.data$schoolid, levels = dec.schoolid )
with(cc.data, boxplot(res1~schoolid))
abline(h=0, col = "red")

ord<-order(unlist(tapply(res1,schoolid,median)))
boxplot(split(res1,schoolid)[ord])
```

```{r}
dec.classid <- as.numeric(names(sort(table(classid))))
cc.data$classid <- ordered(cc.data$classid, levels = dec.classid )
with(cc.data, boxplot(res1~classid))
abline(h=0, col = "red")

```


*construct the residual that utilizes the BLUPs for the random effects.  Do it in these stages:
* i) predict and save zeta0 
* ii) predit and save eta0
* iii) generate a new residual, called resFE_RE which subtracts yhat, zeta0 and eta0 from the outcome
*note: there is an easier way to get the residuals in this case, predict ..., residuals, but we need to do it manually.
```{r}
idx.sch <- match(cc.data$schoolid, sort(unique(cc.data$schoolid)))
idx.cls <- match(cc.data$classid, sort(unique(cc.data$classid))) 
```

```{r}
attach(cc.data)
ranefs <- ranef(fit1)
zeta0 <- ranefs$schoolid[, 1]
eta0 <- ranefs$classid[, 1]
cc.data$zeta0 <- zeta0[idx.sch]
cc.data$eta0 <- eta0[idx.cls]
resFE_RE <- math1st - pred.yhat - cc.data$zeta0 - cc.data$eta0
```


* show that these new residuals, resFE_RE are MUCH LESS (if not completely un-) correlated within school 
* using the same method as before (boxplot?)
### ZARNI
```{r}
with(cc.data, boxplot(resFE_RE~schoolid))
abline(h=1, col="red")

ord2<-order(unlist(tapply(resFE_RE,schoolid,median)))
boxplot(split(resFE_RE,schoolid)[ord2])

```


*examine the two sets of BLUPs (for random effects zeta0 and eta0) for normality
*first 'tag' a single value from each grouping (e.g., school or classroom) so that you only have 
* as many BLUPs as the grouping factor (should be less of an issue in R)
```{r}
plot(density(zeta0))
qqnorm(zeta0)
qqline(zeta0)
```

```{r}
plot(density(eta0))
qqnorm(eta0)
qqline(eta0)
```

*if In a simpler setting, with no classroom level, if H0: \sigma^2_{\zeta_0}=0 were true, 
* and you sampled 100 schools of size 10, what would be a potential expected estimate of \sigma_{\zeta_0} 
* if you know that \sigma_{\epsilon}=1 ?  HINT: the Central Limit Theorem applies.  Simulate to find out.
```{r}

```


*now reload the data and fit a slightly more complicated model:
use "classroom.dta"
gen math1st = mathkind + mathgain
```{r}
dat <- read.csv("classroom.csv")
dat$math1st <- dat$mathkind + dat$mathgain
dat <- dat[(complete.cases(dat)),]
attach(dat)


```



mixed math1st housepov yearstea mathprep mathknow ses sex minority |||schoolid: minority, cov(un) || classid:, reml 
est store M2

```{r}
M2 <- lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (0+minority|schoolid) + (1|schoolid) + (1|classid))
summary(M2)

```
```{r}
idx.sch <- match(dat$schoolid, sort(unique(dat$schoolid)))
idx.cls <- match(dat$classid, sort(unique(dat$classid))) 
```


*manually construct the residual that removes only the 'fixed effects', call this residual: resFE
```{r}
pred.yhat.2 <- predict(M2, re.form = ~0)
ranefs <- ranef(M2)
zeta0 <- ranefs$schoolid[, 1]
dat$zeta0 <- zeta0[idx.sch]
resFE <- dat$math1st - pred.yhat.2 
```

*show that this residual is not independent within schools in some manner.  Hint: boxplot?
```{r}



```

*construct the residual that utilizes the BLUPs for the random effects.  Do it in these stages:
* i) predict and save zeta0 AND zeta1 (you need to give them in reverse order in STATA - ask me why if you want)
* ii) predit and save eta0
* iii) generate a new residual, called resFE_RE which subtracts yhat, zeta0, MINORITY*zeta1 and eta0 
```{r}
ranerr <- ranef(M2)
zeta0 <- ranerr$schoolid[,1]
zeta1 <- ranerr$schoolid[,2]
eta0 <- ranerr$classid[,1]
dat$zeta0 <- dat$zeta0[idx.sch]
dat$zeta1 <- dat$zeta1[idx.sch]
dat$eta0 <- dat$eta0[idx.cls]
resFE_RE <- dat$math1st - (pred.yhat.2 + dat$zeta0 + dat$minority * dat$zeta1 + dat$eta0)
```


*      from the outcome

* show that these new residuals, resFE_RE are MUCH LESS (if not completely un-) correlated within school 
* using the same method as before (boxplot?)
```{r}



```

*examine the three sets of BLUPs (for random effects zeta0 and eta0) for normality
```{r}
qqnorm(zeta0)
qqline(zeta0)
```

```{r}
qqnorm(zeta1)
qqline(zeta1)
```

```{r}
qqnorm(eta0)
qqline(eta0)
```


*first 'tag' a single value from each grouping (e.g., school or classroom) so that you only have 
* as many BLUPs as the grouping factor (should be less of an issue in R)

*plot zeta0 vs. zeta1 to see whether the estimated correlation is consistent with the observed. 
* Use tag to subset as before

#No Correlation
```{r}
library(calibrate)
```


```{r}
plot(jitter(zeta0), zeta1)
text(zeta0, zeta1, labels=schoolid, cex= 0.7, pos=3)

plot(jitter(zeta0[1:10], zeta1[1:10]))
textxy(zeta0[1:10], zeta1[1:10], schoolid[1:10])

```


*track down those odd points in the scatterplot.  What schools are they?
```{r}



```
